Hippocampal segmentation with MAGeT

### 2013-02-05 paper.Rnw
Complete introduction with rewrites to include even more literature

%  Automated methods 
%
%  Heckemann2006 - 
%  Heckemann2011 -  Automatic morphometry in AD and MCI 
%    - (MAPER). multi-atlas. 30 atlases, entire ADNI screening  
%  Collins2010 - .. seg of HC and AM by augmenting ANIMAL
%    - multi-atlas 80 atlases, atlas selection (NMI)
%  Aljabar2009 - atlas selection and effect on accuracy
%    - 
%  Leung2010 - X-sectional and longitudinal HC vol in MCI, AD
%    - (MAPS), multi-atlas + STAPLE. 55 atlases. baseline+1yr of ADNI seg.
%    - atrophy rates
%  Lotjonen2010 - Fast and robust
%    - IBSR/ADNI.  
%  Mouiha2011b - Hippocampal atrophy rates in AD
%    - compare SNT vs FreeSurfer on ADNI
%
Fully-automated segmentation techniques overcome the need for human expertise.
In particular, {\it atlas-based} approaches make use of one or more expertly
segmented MRIs (the atlas library) as templates to guide the segmentation of a
query subject's MRI
\cite{Aljabar2009,Collins2010,Heckemann2011,Leung2010,Lotjonen2010,Leung2010,Wolz2010}.
The basic strategy is as follows. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration techniques
\citep{Collins1995,Klein2009}.  Atlas labels are then transformed by this
warping to the subject's space, and a {\it label fusion} technique, such as
voxel-wise voting, is used to merge the competing label definitions from each
atlas into a final segmentation for the subject.

Multi-atlas hippocampal segmentation algorithms achieve some of the best
automated segmentation accuracies to-date, and approach the performance of a
manual rater
\citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010}, but rely
on an atlas library containing anywhere between 30 and 80 expertly labeled
images to achieve this accuracy.  As already noted, the production of an atlas
library requires significant manual effort and expertise, as does the
modification of an existing atlas library in order to better suit a target
population or investigate a specific research hypothesis.  Therefore, in this
paper we tackle the problem of producing accurate segmentations using small
numbers of manually segmented atlases.

In this paper we propose an extension to the basic multi-atlas-based
segmentation scheme to address this issue.  Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the subjects in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion.  We
call our approach MAGeT brain: {\em m}ultiple {\em a}utomatically {\em
ge}nerated {\em t}emplate {\em brain} segmentation. The intuition driving this
approach is that by generating a template library we leverage the unique
neuroanatomy of subject population on hand to initialize the segmentation
process and improve accuracy over direct propagation from the atlas library to
unlabelled subjects while also using fewer manually segmented atlases.

The insight of generating a template library is not new.  Heckemann
\citep{Heckemann2006} briefly explored "indirect" segmentation by taking a
single atlas and propagating those labels to intermediate subjects before fusing
them in a target image space.  Heckemann compared indirect single-atlas
segmentation to direct multi-atlas segmentation found the indirect approach
performed worse.  In this paper we continue the same line of investigation but
explore the performance when using multiple atlases, as well as several
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010} is another example of indirect segmentation
previously explored.  LEAP proceeds by iteratively segmenting unlabelled images
most similar to the atlas library images and then incorporating the
now-labelled images into the atlas library for future iterations.  30 manually
labelled atlases are required to start the process.  The novelty explored in
our current work is demonstrate the viability of acheiving comparable
segmentation accuracy while using significantly fewer manually created atlases.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion\citep{Collins2010}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follows.  An atlas library and unlabelled MR images are given as input.  Each
atlas image is nonlinearly registered to each unlabelled image, and then each
atlas' labels are propagated via the resulting transformations.  The resulting
labels are fused to produce a single, definitive segmentation by some label
fusion method (for exampl, by voxel-wise majority voting).

>>>>

Fully-automated segmentation techniques require no human expertise beyond that
of initial setup or training. One broad class of automated approach is that of
the {\em model-based} approaches which employ models of anatomical feature
variation to constrain segmentations. In this paper we focus on a subclass known
as {\em multi-atlas} approaches which make use of a series of expertly segmented
MRIs (the atlas library) as models to guide the segmentation of a query
subject's MRI, and achieve some of the best automated hippocampal segmentation
accuracies to-date.

The multi-atlas segmentation technique was pioneered by Heckemann et al. with
the MAPER algorithm \citep{Heckemann2006,Heckemann2011}. The basic strategy is as
follows. Each atlas is warped to fit a subject's neuroanatomy using nonlinear
registration techniques \citep{Collins1995,Klein2009}. Atlas labels are then
transformed by this warping to the subject space and a {\it label fusion}
technique, such as voxel-wise voting, is used to merge the label definitions
from each atlas into a definitive segmentation. Typically, an atlas selection
approach is used to improve accuracy by only using labels from atlases that are
most similar to a subject image (cross-correlation of image intensities is a
common similarity metric) \citep{Aljabar2009}.

% - Warfield2004: STAPLE as an alternative decision fusion technique?

Multi-atlas techniques have been applied very successfully to hippocampal
segmentation. Collins et al. found that with an atlas library of 80 atlas
images, the use of the ANIMAL nonlinear registration algorithm, the normalised
mutual information as similarity metric for atlas selection, and majority vote
for decision fusion they were able to achieve a mean Dice Kappa similarity score
(DSC) to manual labels of 0.886\citep{Collins2010}. The Alzheimer's Disease
Neuroimaging Initiative (ADNI) is a commonly used benchmarking dataset of MR
images of controls and patients with MCI or Alzheimer's (see Methods for more
information on the ADNI dataset).  Leung et al. tuned a multi-atlas  approach to
the segmentation of ADNI1 1 year dataset of images and using 55 atlases and the
STAPLE decision fusion method \citep{Warfield2004} achieved an mean DSC of 0.89
to the manual segmentations supplied with the ADNI data  \citep{Leung2010}. The
MAPER whole brain segmentation algorithm (described above), using 30 atlases,
achieved a mean DSC score of 0.889 on all ADNI1 baseline images
\citep{Heckemann2011}. In \citep{Lotjonen2012}, the authors use a proprietary
non-linear registration method based on intensity differences, and
post-processing step taking into account these intensity differences to tidy up
fused segmentations. Using images from the ADNI1 baseline dataset, they achieved
a mean DSC of 0.885 with 30 atlases. With the LEAP algorithm, Wolz et al.
explored an elegant modification to the basic strategy: the atlas library is
grown, beginning with a set of manually labelled atlases and successively
incorporating unlabelled subject images after being labelling using multi-atlas
techniques\citep{Wolz2010}.  The sequence in which subject images are labelled
is chosen so that the similarity between the atlas images and the target images
is minimised at each step, effectively allowing for deformations between very
dissimilar images to be broken up into sequences of smaller deformations.  With
an atlas library of 30 MR images, LEAP was used to segment the ADNI1 baseline
dataset, achieving a mean Dice score of 0.85 with manual segmentations.

While not purely nulti-atlas techniques, there are several important algorithms
for hippocampal segmentation that inform our approach. The popular FreeSurfer
application's whole brain segmentation algoritm uses a probabilistic atlas of
anatomical and tissue classes along with spatial constraints for class labels
encoded using a Markov random field model \citep{Fischl2002}. When segmenting
hippocampal subfields, FreeSurfer employs a Bayesian inference algorithm using a
probabilistic atlas of anatomical classes as a prior, and a likelihood model of
how those classes translate into MR image intensities, both trained on manual
segmentations of high resolution MR images \citep{Fischl2009}. Yushkevitch et
al. describe a semi-automated method for hippocampal subfield segmentation of
focal T2 images\citep{Yushkevitch2010}.  The unlabelled MR image must be
manually partitioned into 'head', 'body' and 'tail', and then multi-atlas
methods are used to segment the image.  Finally, an AdaBoost-based bias
correction classifier is trained on texture, spatial location, and intensities
of manual segmentations and is applied to fix mislabelled voxels.  
%  Chupin's use of shape-based priors 
%   - Mouiha2011b - Hippocampal atrophy rates in AD: compare SNT vs FreeSurfer
%     on ADNI
%

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
subject image by voxel intensity profile segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a role\citep{Aljabar2009}.
Carmichael et al. explored this directly and found that when using only one
atlas the important factors leading to improved accuracy are that the atlas have
neuroanatomical features that match the subject, and that the atlas segmentation
use the same protocol as the gold-standard \citep{Carmichael2005}. Nestor et al.
found that hippocampal segmentation protocols that include more dorsal
white-matter and posterior anatomy tended to produce higher overlap and better
accuracy at distinguishing disease classes in the ADNI1 1 year dataset
\citep{Nestor2012}. These results suggest both atlas library neuroanatomy and
delineation protocol play a significant role in the resulting segmentation.

Considered along with our earlier discussion on the difficulty of producing
manual segmentations of MR images and the need for adaptable segmentation
definitions in order to conduct research, this presents a real problem of labour
and expertise when using existing multi-atlas segmentation methods which rely on
relatively large atlas libraries (typically between 30 and 80 atlases). Indeed,
it may be especially prohibitive to use these methods in situations where
producing a single atlas is challenging (e.g. histology-based atlases, or
atlases from very high resolution images). In this paper we address the problem
of producing accurate segmentations using small numbers of manually segmented
atlases.

% MAGeT bootstrapping: tune the template library to your dataset, reduce the
% number of input atlases you need, with the upshot that with fewer atlases it's
% that much easier to change definitions in order to test hypothesis

Our algorith, called MAGeT brain ({\em M}ultiple {\em A}utomatically {\em
Ge}nerated {\em T}emplates), is an extension to the basic multi-atlas-based
segmentation schema\citep{Chakravarty2012}. Principally, we explore the
possibility of using a small atlas library to bootstrap a much larger
\emph{template library} composed of images taken from the target population.
The template library is then used to segment the subjects in a similar fashion
to basic multi-atlas segmentation: by label propagation and label fusion. The
intuition driving this approach is that by generating a template library we
leverage the unique neuroanatomy of subject population on hand to initialize the
segmentation process and improve accuracy over direct propagation from the atlas
library to unlabelled subjects while also using fewer manually segmented
atlases.

The insight of generating a template library is not new.  Heckemann et al.
compared "indirect" segmentation -- taking a single atlas and propagating the
labels to intermediate subjects before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is demonstrate the viability of
achieving comparable segmentation accuracy while using significantly fewer
manually created atlases.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion\citep{Collins2010}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follows.  An atlas library and unlabelled MR images are given as input.  Each
atlas image is nonlinearly registered to each unlabelled image, and then each
atlas' labels are propagated via the resulting transformations.  The resulting
labels are fused to produce a single, definitive segmentation by some label
fusion method (for example, by voxel-wise majority voting).
